### 2017-12-12 submission/manuscript.Rmd
tone down conclusions
fix grammar and structure of methods and declaration section

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma. Overall, our results support a framework by which early localized community changes give rise to polyp formation. With the host as a potential catalyst, new niches arise by which non-resident CRC-associated microbes can then gain a foothold and create an environment that allows more of these microbes to colonize. This exacerbates the existing community changes and creates a vicious cycle conducive of carcinoma formation. Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Although there are power limitations associated with adenoma, this report highlights the strong influence the microbiota has on CRC development.

@@

***Obtaining Data Sets:***
Studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016]. Additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature. Studies that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis were included. Studies using either 454 or Illumina sequencing technology were included. Only data sets that had sequences available for analysis were included. Some studies did not have publically available sequences or did not have metadata in which the authors were able to share. After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016]. Additionally, after sequence processing all the case samples for the Kostic study only had 100 sequences remaining and was not used. This left a total of 14 studies for which analysis could be completed.

@@

***Sequence Processing:***
For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults, the stated quality cut-offs were used instead. Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many with and without disease were above and below the overall median value within the specific study. Relative risk significance testing utilized the chi-squred test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Differences between the obtained AUC versus a random model AUC was assessed using T-tests. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched) tissue samples. Where applicable for each study, predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarily, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.

@@

***Assessing Important Random Forest Model Variables:***
The genus based models collected the top 10 most important variables from each training set and assessed how many times that genera showed up in the top 10 across each study. The OTU based models recorded the medians for each OTU across 100 different 80/20 splits of the data for each study. The lowest classification for each OTU was obtained using the RDP database and the number of times the specific classification occured in the top 10 across studies was recorded. For the adenoma tissue genus and OTU models there was only one matched and unmatched study and these results were grouped together for the counting of the top 10.

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless mentioned otherwise, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

@@

The authors would like to thank all the study participants who were a part of each of the individual studies utilized. We would also like to thank each of the study authors for making their data available for use. Finally we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.

>>>>

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma. Overall, our results support a framework by which early localized community changes give rise to polyp formation. With the host as a potential catalyst, new niches arise by which non-resident CRC-associated microbes can then gain a foothold and create an environment that allows more of these microbes to colonize. This exacerbates the existing community changes and creates a vicious cycle conducive of carcinoma formation. Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Although there are power limitations associated with adenoma, this report highlights the strong associations the microbiota has on CRC.

@@

***Obtaining Data Sets:***
Studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016]. Additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature. Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included. Only data sets that had sequences available for analysis were included. Some studies did not have publically available sequences or did not have metadata in which the authors were able to share and were excluded. After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016]. Additionally, after sequence processing all the case samples for the Kostic study only had 100 or less sequences remaining and was not used. This left a total of 14 studies for which analysis could be completed.

@@

***Sequence Processing:***
For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults, the stated quality cut-offs were used instead. Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and variable region effects. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many with and without disease were above and below the overall median value within the specific study. Relative risk significance testing utilized the chi-squred test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each individual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. Finally, the power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarily, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.

@@

***Assessing Important Random Forest Model Variables:***
The genus based models collected the top 10 most important variables, measured by Mean Decrease in Accuracy (MDA), from each training set and assessed how many times that genera showed up in the top 10 across each study. The OTU based models recorded the medians for each OTU across 100 different 80/20 splits of the data for each study. The lowest classification for each OTU was obtained using the RDP database and the number of times the specific classification occured in the top 10 across studies was recorded. For the adenoma tissue genus and OTU models there was only one matched and unmatched study and these results were grouped together for the counting of the top 10.

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless otherwise mentioned, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

@@

The authors would like to thank all the study participants who were a part of each of the individual studies utilized. We would also like to thank each of the study authors for making their data available for use. Finally, we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.

### 2017-12-12 submission/manuscript.tex
tone down conclusions
fix grammar and structure of methods and declaration section

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 16S rRNA gene sequencing analysis were included. Studies
using either 454 or Illumina sequencing technology were included. Only
data sets that had sequences available for analysis were included. Some
studies did not have publically available sequences or did not have
metadata in which the authors were able to share. After these filtering
steps, the following studies remained: Ahn, \emph{et al.} {[}26{]},
Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]}, Burns,
\emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea, \emph{et
al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et al.}
{[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.} {[}15{]},
Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang,
\emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]}, and Zeller,
\emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not included
becasue the 90 individuals analyzed within the study are contained
within the larger Baxter study {[}16{]}. Additionally, after sequence
processing all the case samples for the Kostic study only had 100
sequences remaining and was not used. This left a total of 14 studies
for which analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the by searching the respective accession
number of the study following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}26{]} and for four studies was obtained directly from
the authors {[}12,18,23,25{]}. Each study was processed using the mothur
(v1.39.3) software program {[}32{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity using the OptiClust algorithm {[}34{]} was utilized.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.2) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study
and variable region effects in feces and study, variable region, and
individual effects in tissue. Relative risk was analyzed using both the
epiR (v0.9.87) and metafor (v2.0.0) packages {[}39,40{]} by assessing
how many with and without disease were above and below the overall
median value within the specific study. Relative risk significance
testing utilized the chi-squred test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}42,43{]}. Differences between the obtained AUC versus a random model
AUC was assessed using T-tests. Power analysis and estimations were made
using the pwr (v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All
figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
packages {[}46,47{]}.

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each inidividual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched) tissue samples. Where applicable for each study,
predictions for adenoma and carcinoma were tested. This same approach
was then applied at the OTU level with the exception that instead of
testing on the other studies, a 10-fold cross validation was utilized
and 100 different models were created based on random 80/20 splitting of
the data to generate a range of expected AUCs. For OTU based models, the
CRC associated genera included all OTUs that had a taxonomic
classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
study was assessed for an effect size ranging from 1\% to 30\%. An
estimated sample n for these effect sizes was also generated based on
80\% power. For comparisons in which normal versus adenoma were made the
carcinoma samples were excluded from each respective study. Similarily,
for comparisons in which normal versus carcinoma were made the adenoma
samples were excluded from each respective study.

@@

\textbf{\emph{Assessing Important Random Forest Model Variables:}} The
genus based models collected the top 10 most important variables from
each training set and assessed how many times that genera showed up in
the top 10 across each study. The OTU based models recorded the medians
for each OTU across 100 different 80/20 splits of the data for each
study. The lowest classification for each OTU was obtained using the RDP
database and the number of times the specific classification occured in
the top 10 across studies was recorded. For the adenoma tissue genus and
OTU models there was only one matched and unmatched study and these
results were grouped together for the counting of the top 10.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless mentioned otherwise, the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.

>>>>

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 454 or Illumina 16S rRNA gene sequencing analysis were
included. Only data sets that had sequences available for analysis were
included. Some studies did not have publically available sequences or
did not have metadata in which the authors were able to share and were
excluded. After these filtering steps, the following studies remained:
Ahn, \emph{et al.} {[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim,
\emph{et al.} {[}22{]}, Burns, \emph{et al.} {[}27{]}, Chen, \emph{et
al.} {[}19{]}, Dejea, \emph{et al.} {[}24{]}, Flemer, \emph{et al.}
{[}12{]}, Geng, \emph{et al.} {[}28{]}, Hale, \emph{et al.} {[}18{]},
Kostic, \emph{et al.} {[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy,
\emph{et al.} {[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et
al.} {[}23{]}, and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]}
study was not included becasue the 90 individuals analyzed within the
study are contained within the larger Baxter study {[}16{]}.
Additionally, after sequence processing all the case samples for the
Kostic study only had 100 or less sequences remaining and was not used.
This left a total of 14 studies for which analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}26{]} and for four studies was obtained directly from
the authors {[}12,18,23,25{]}. Each study was processed using the mothur
(v1.39.3) software program {[}32{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}34{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.2) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
repeat sampling of individuals (tissue only), and variable region
effects. Relative risk was analyzed using both the epiR (v0.9.87) and
metafor (v2.0.0) packages {[}39,40{]} by assessing how many with and
without disease were above and below the overall median value within the
specific study. Relative risk significance testing utilized the
chi-squred test. \(\beta\)-diversity differences utilized a Bray-Curtis
distance matrix and PERMANOVA executed with the vegan (v2.4.4) package
{[}41{]}. Random Forest models were built using both the caret (v6.0.77)
and randomForest (v4.6.12) packages {[}42,43{]}. Power analysis and
estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
packages {[}44,45{]}. All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages {[}46,47{]}.

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each individual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched). Where applicable for each study, predictions
for adenoma and carcinoma were then tested for feces, matched tissue,
and unmatched tissue. This same approach was then applied at the OTU
level with the exception that instead of testing on the other studies, a
10-fold cross validation was utilized and 100 different models were
created based on random 80/20 splitting of the data to generate a range
of expected AUCs. For OTU based models, the CRC associated genera
included all OTUs that had a taxonomic classification to
\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
\emph{Porphyromonas}. Finally, the power of each study was assessed for
an effect size ranging from 1\% to 30\%. An estimated sample n for these
effect sizes was also generated based on 80\% power. For comparisons in
which normal versus adenoma were made the carcinoma samples were
excluded from each respective study. Similarily, for comparisons in
which normal versus carcinoma were made the adenoma samples were
excluded from each respective study.

@@

\textbf{\emph{Assessing Important Random Forest Model Variables:}} The
genus based models collected the top 10 most important variables,
measured by Mean Decrease in Accuracy (MDA), from each training set and
assessed how many times that genera showed up in the top 10 across each
study. The OTU based models recorded the medians for each OTU across 100
different 80/20 splits of the data for each study. The lowest
classification for each OTU was obtained using the RDP database and the
number of times the specific classification occured in the top 10 across
studies was recorded. For the adenoma tissue genus and OTU models there
was only one matched and unmatched study and these results were grouped
together for the counting of the top 10.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless otherwise mentioned, the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.
